Supplementary material for  \nBayesian multi-proxy reconstruction of the early Eocene latitudinal temperature gradient

### 2023-02-07 manuscript/manuscript.Rmd
add R2 modern, discussion first part

The early Eocene is the focus of a substantial body of palaeoclimate literature [@Berggren1998; @Thomas2000; @Hyland2017; @Lunt2021], as it is the interval with the warmest sustained temperatures of the Cenozoic [@Pross2012], and potentially represents an analogue for extreme climate warming scenarios [@Burke2018]. For palaeoclimate modelers, the early Eocene presents a challenge: Geochemical proxy data [@Hollis2019] and ecological data [@Marwick1994; @Greenwood1995] indicate tropical to subtropical conditions up to high latitudes, but early climate models have not been able to model warm polar temperatures and a shallow latitudinal temperature gradient under realistic conditions [@Sloan1990; ...]. Although some more recent climate models have been able to better align their simulations with the proxy record [@Huber2013;...], the nature of the early Eocene temperature gradient remains a matter of debate [@Tierney2017]. The shape and magnitude of the inferred EECO latitudinal gradient depends on the selection of proxies, on the method for dealing with spatial gaps and uneven sampling, on the inclusion or exclusion of terrestrial, shallow and deep marine data, while differential reporting of the results complicates comparisons of different approaches [@Evans2018; @Hollis2019; @Inglis2020]. Our goal is to provide a robust and relatively simple method for inferring and comparing temperature gradients that can be readily used with different proxy types and with patchy records. We use this model to quantify the latitudinal SST gradient and the global mean SST of the early Eocene, but the unspecific design of our method means it can be readily applied to other time intervals and proxy records.

@@

@@

```
## Model validation
<!-- - Show that the 4-parameter logistic function can adequatly describe the modern temperature gradient
- Show that the modern gradient can be reproduced with Eocene sampling distribution
-->
[Fig 3 - Modern gradient]
The average modern temperature gradient can be closely approximated with our model when using the full modern SST dataset (Fig. 3a). Reducing the amount and the spatial coverage of data points to match the early Eocene sampling distribution ($n = 34$) still results in a close match of the modeled gradient with the empirical gradient (Fig. 3b). The 95% credible interval (CI) of reconstructed gradients (Fig. 3b) is noticeably narrower than the spread of empirical temperature values (Fig. 3a). The modern, latitudinal gradient, modeled with all data, spans `r modern_gradient$"50%"` degC from the equator to the poles. When modeled with only the early Eocene sampling locations, the median temeprature gradient is `r modern_eocene_gradient$"50%"`  (95% CI: `r modern_eocene_gradient$"2.5%"` - `r modern_eocene_gradient$"97.5%"`) degC. A random sampling of modern temperatures at latitudes corresponding to the sampled Eocene palaeolatitudes would thus allow the reconstruction of the modern latitudinal temperature gradient with good accuracy, with expected deviations from the actual gradient as low as `r modern_eocene_gradient_diff$"2.5%"` to `r modern_eocene_gradient_diff$"97.5%"` degC (95 % CI).

Additional model runs with simulated SST data show that latitudinal gradients following a flat, linear or quadratic function can also be adequately reconstructed by our model (Fig. S4). The generalised logistic function underlying the reconstruction leads to the inferred gradient being pushed towards a sigmoidal shape in the absence of proxies (see very high latitudes in Fig S4a,c). This is intended, as in the absence of data, the shape of the reconstructed gradient should be broadly similar to the sigmoidal shape of the modern SST gradient, as it is the only empirically observed SST gradient.

@@

# Discussion

>>>>

The early Eocene is the focus of a substantial body of palaeoclimate literature [@Berggren1998; @Thomas2000; @Hyland2017; @Lunt2021], as it is the interval with the warmest sustained temperatures of the Cenozoic [@Pross2012], and potentially represents an analogue for extreme climate warming scenarios [@Burke2018]. For palaeoclimate modelers, the early Eocene presents a challenge: Geochemical proxy data [@Hollis2019] and ecological data [@Marwick1994; @Greenwood1995] indicate tropical to subtropical conditions up to high latitudes, but early climate models have not been able to model warm polar temperatures and a shallow latitudinal temperature gradient under realistic conditions [@Sloan1990; ...]. Although some more recent climate models have been able to better align their simulations with the proxy record [@Huber2013;...], the nature of the early Eocene temperature gradient remains a matter of debate [@Tierney2017]. The shape and magnitude of the inferred EECO latitudinal gradient depends on the selection of proxies, on the method for dealing with spatial gaps and uneven sampling, on the inclusion or exclusion of terrestrial, shallow and deep marine data, while differential reporting of the results complicates comparisons of different approaches [@Evans2018; @Hollis2019; @Inglis2020]. Our goal is to provide a robust, yet simple method for inferring and comparing temperature gradients that can be readily used with different proxy types and with patchy records. We use this model to quantify the latitudinal SST gradient and the global mean SST of the early Eocene, but the unspecific design of our method means it can be readily applied to other time intervals and proxy records.

@@

# modern R2
R2_modern <- readRDS("results/modern/modern_model_empirical_R2.rds")

@@

```
## Model validation
<!-- - Show that the 4-parameter logistic function can adequatly describe the modern temperature gradient
- Show that the modern gradient can be reproduced with Eocene sampling distribution
-->
[Fig 3 - Modern gradient]
The average modern temperature gradient can be closely approximated with our model when using the full modern SST dataset (Fig. 3a); almost all of the variation in the empirical median temperatures in bins of 1 deg absolute latitude is explained by the modeled gradient ($R^2 = `r R2_modern`$). Reducing the amount and the spatial coverage of data points to match the early Eocene sampling distribution ($n = 34$) still results in a close match of the modeled gradient with the empirical gradient (Fig. 3b). The 95% credible interval (CI) of reconstructed gradients (Fig. 3b) is noticeably narrower than the spread of empirical temperature values (Fig. 3a). The modern, latitudinal gradient, modeled with all data, spans `r modern_gradient$"50%"` degC from the equator to the poles. When modeled with only the early Eocene sampling locations, the median temeprature gradient is `r modern_eocene_gradient$"50%"`  (95% CI: `r modern_eocene_gradient$"2.5%"` - `r modern_eocene_gradient$"97.5%"`) degC. A random sampling of modern temperatures at latitudes corresponding to the sampled Eocene palaeolatitudes would thus allow the reconstruction of the modern latitudinal temperature gradient with good accuracy, with expected deviations from the actual gradient as low as `r modern_eocene_gradient_diff$"2.5%"` to `r modern_eocene_gradient_diff$"97.5%"` degC (95 % CI).

Additional model runs with simulated SST data show that latitudinal gradients following a flat, linear or quadratic function can also be adequately reconstructed by our model (Fig. S4). The generalised logistic function underlying the reconstruction leads to the inferred gradient being pushed towards a sigmoidal shape in the absence of proxies (see very high latitudes in Fig. S4a,c). This is intended, as in the absence of data, the shape of the reconstructed gradient should be broadly similar to the sigmoidal shape of the modern SST gradient, as it is the only empirically observed SST gradient.

@@

# Discussion
The Bayesian model presented herein accurately reconstructs the modern, latitudinal SST gradient from a patchy and relatively small (n = 34) sample of temperature data. This is an advancement over previously used linear or quadratic approximations [e.g. @Refs], which are inherently unable to capture the sigmoidal shape of the modern SST gradient, although our generalised logistic function is flexible enough to approximate those shapes, too. As such, our model presents an alternative to non-parametric methods for inferring latitudinal temperature gradients. Non-parametric models can very flexibly follow the shape of an unknown temperature gradient [e.g. @Zhang2019; @Jones2022], but have the disadvantage of having no model parameters that can be readily compared between data sets. In contrast, the four parameters that describe the shape of our generalised logistic model clearly relate to the magnitude, steepness and the latitude of greatest steepness of the modeled gradient. Given that at least vague prior information on the shape of latitudinal temperature gradients on Earth should exist for any geological time period, we argue that parametric methods should be preferred for reconstructing those gradients.

The ease with which such prior information can be integrated is a major advantage of our method: The Bayesian framework allows to specify arbitrary priors on the four parameters that control the shape of the gradient (see Methods). A further improvement to existing methods is that the hierarchical structure of our model explicitly accounts for variation of temperature estimates at individual localities, and for uncertainty associated with the ecological temperature proxies. An extension of the model additionally accounts for measurement uncertainty of the geochemical proxies (Supplementary Materials).

### 2023-02-07 manuscript/supplementary_material.Rmd
add R2 modern, discussion first part

```{r out.width='100%', echo = FALSE, fig.cap = "Fig S2. Latitudinal temperature gradient with just the geochemical proxies (orange), showing the median (line) and 95% credibal interval (shading). Symbols with vertical lines show the median and 95% credible intervals of $\\mu$. The grey line and shading in the background show the latitudinal temperature gradient with the geochemical and ecological proxy data."}
setwd('..') # move up to parent folder of /manuscript
knitr::include_graphics("D://Github/PalaeoClimateGradient/figures/SM/FigS2_geochem_eeco.png", error = FALSE)
```

@@

>>>>

```{r out.width='100%', echo = FALSE, fig.cap = "Fig S2. Latitudinal temperature gradient with just the geochemical proxies (orange), showing the median (line) and 95% credibal interval (shading). Symbols with vertical lines show the median and 95% credible intervals of $\\mu$. The blue line and shading in the background show the latitudinal temperature gradient with the geochemical and ecological proxy data."}
setwd('..') # move up to parent folder of /manuscript
knitr::include_graphics("D://Github/PalaeoClimateGradient/figures/SM/FigS2_geochem_eeco.png", error = FALSE)
```

@@

## S4. Different shapes of temperature gradients

```{r out.width='100%', echo = FALSE, fig.cap = "Fig S4. Simulated latitudinal temperature gradients sampled at the absolute latitudes of the EECO data. The dark grey points show the simulated temperature data, the black line shows the shape of the true gradient, and the medians (lines) and 95% credibal intervals (shadings) of the inferred gradients using the model are shown in red. \\\n a) linear gradient; $SST = 30-|latitude|/3$ \\\n b) flat gradient; $SST = 25$ \\\n c) quadratic gradient; $SST = 30-|latitude|^2/270$ \\\n d) logit-shaped gradient; $SST = 14-3*log(|latitude|/(90-|latitude|))$ \\\n  Except for d), which is very unlike the modern gradient or reconstructed gradients, all latitudinal temperature gradients are well captured by the model in the latitudinal range for which data is provided."}
setwd('..') # move up to parent folder of /manuscript
knitr::include_graphics("D://Github/PalaeoClimateGradient/figures/SM/FigS4_different_gradients.png", error = FALSE)
```
